Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-12 paper.Rnw
Several fixes

@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\textbf{Results:} \textbf{[TODO: update with FEP-XVAL and subfield XVAL results.]} 
Using 9 atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of \Sexpr{mb.kappa.best} 
(to \psegs{}) over 10-folds of Monte Carlo 
significantly lower variability in DSC than \multiatlas{} segmentation.
In a LOOCV, \mb{} accurately reproduces the subfields... .

\mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt{})
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.\\ \textbf{Conclusion:} We demonstrate
that \mb{} produces accurate hippocampal segmentations using only 5 atlases over
different hippocampal definitions, disease populations, and acquisition types,
as well as showing that accurate identification of the hippocampal subfields is
possible.
\end{abstract}

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion. The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation agreement, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation. Given
these findings, in the remainder of our experiments only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

With at least five templates, \mb{} shows a mean improvement in similarity score
over \multiatlas{} segmentation when using the same size of atlas library and
majority vote label fusion (Figure \ref{fig:ADNI1-xval-k-diff}). The magnitude
of improvement increases with the size of template library, and shows
diminishing returns with larger atlas libraries. Peak improvement
(+\Sexpr{round(mb.diff.best, digits=3)} DSC) is found with a single atlas
and template library of 19 images.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expadni{}: \expadnidesc{}}
\label{exp:adni}
To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to a large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated whole
hippocampal segmentations (\snt{}) provided by ADNI.

@@

\paragraph{\adnidataset{} dataset}  %%%%%%%%%%%
The {\em \adnidataset{}} standardized dataset contains $1919$ images of subjects
diagnosed with Alzheimer's disease (AD), late-onset mild cognitive impairment
(LMCI), or who are identified as cognitively normal (CN). See Section
\ref{adnidataset} for study details, inclusion criteria and imaging
characteristics.

@@

Figure \ref{fig:WAxval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subfield across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subfields are shown in Table
\ref{tab:WAxval-perturb-kappa}. We find that the CA4/DG subfield shows the highest mean DSC score of
\Sexpr{mean(CA4DG)}$\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subfields having scores of \Sexpr{mean(CA1)}$\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)}$\pm$\Sexpr{sd(Sub)}, respectively. Both the CA4/DG and
molecular regions score below 0.5. These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion).
The whole hippocampus is segmented with a mean DSC score
of\Sexpr{mean(Whole)}$\pm$\Sexpr{sd(Whole)}. Table
\ref{tab:WAxval-perturb-kappa} also shows mean DSC overlap scores for simulated
manual segmentations with error introduced by a one-voxel translation in all
directions. For all subfields, \mb{} DSC scores are well within the error bounds
set by these segmentations.

@@

>>>>

# http://druedin.com/2013/02/16/p-level-in-sweave/
plevel <- function (x, strict=FALSE) {
  # levels of p-values, for Sweave
  # strict cuts at 0.05, otherwise cuts at 0.1
  if (x>0.1 & strict==FALSE) p <- "p>0.1"  # not significant
  if (x>0.1 & strict==TRUE)  p <- "p>0.05" # not significant
  if (x<=0.1 & strict==FALSE) p <- "p<0.1"  # significant
  if (x<=0.1 & strict==TRUE)  p <- "p>0.05" # not significant
  if (x<=0.01)  p <- "p<0.01"  # significant
  if (x<=0.05)  p <- "p<0.05"  # significant
  if (x<=0.001) p <- "p<0.001" # significant
  return(p)
  }
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\textbf{Results:} 
ADNI-XVAL
In the ADNI cross-validation, using 9 atlases \mb{} achieves a mean Dice's
Similarity Coefficient (DSC) score of \Sexpr{mb.kappa.best} with respect to
manual whole hippocampus segmentations, and also exhibits significantly lower
variability in DSC scores than \multiatlas{} segmentation.  In the younger,
psychosis dataset, \mb{} achieves a mean DSC score of \Sexpr{fep.mb.kappa.best}
and produces volumes which agree with manual segmentation volumes better than
those produced by the \freesurfer and \fsl methods (mean difference in volume:
$80mm^3$, $1600mm^3$, and $800mm^3$, respectively). Across the \adnidataset
dataset,

ADNI-1YR
Subfields

In a LOOCV, \mb{} accurately reproduces the subfields... .

\mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt{})
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.

\textbf{Conclusion:} We demonstrate that \mb{} produces accurate hippocampal
segmentations using only 5 atlases over different hippocampal definitions,
disease populations, and acquisition types, as well as showing that accurate
identification of the hippocampal subfields is possible.
\end{abstract}

@@

<<ADNI1-xval-scores,include=FALSE>>=
k.1t = round(mean(subset(jens_xval_mean, 
                           reg_method   == "ANTS" & 
                           method.mb    == "Majority Vote" &
                           templates.mb == 1 & 
                           atlases      == 9)$k.mb), digits=3)
k.5t = round(mean(subset(jens_xval_mean, 
                           reg_method   == "ANTS" & 
                           method.mb    == "Majority Vote" &
                           templates.mb == 5 & 
                           atlases      == 9)$k.mb), digits=3)
k.9t = round(mean(subset(jens_xval_mean, 
                           reg_method   == "ANTS" & 
                           method.mb    == "Majority Vote" &
                           templates.mb == 9 & 
                           atlases      == 9)$k.mb), digits=3)
k.17t = round(mean(subset(jens_xval_mean, 
                           reg_method   == "ANTS" & 
                           method.mb    == "Majority Vote" &
                           templates.mb == 17 & 
                           atlases      == 9)$k.mb), digits=3)
k.9t.animal = round(mean(subset(jens_xval_mean, 
                                 reg_method   == "ANIMAL" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 9 & 
                                 atlases      == 9)$k.mb), digits=3)

# Does weighted voting significantly improve scores? 
# Let's correlate DSC scores at (a, t) = (1, 5) and (1, 15), (9, 5) and (9, 15) 
ants   = subset(jens_xval_mean, reg_method=='ANTS')
animal = subset(jens_xval_mean, reg_method=='ANIMAL')
mavote = subset(ants, method.mb == 'Majority Vote')
mivote = subset(ants, method.mb == 'NMI Vote')
ants.stats = ddply(ants, c("atlases", "templates.mb"), 
                   function (df) {
                     test = cor.test(subset(df, method.mb == 'Majority Vote')$k.mb, 
                                     subset(df, method.mb == 'NMI Vote')$k.mb)
                     data.frame(
                       estimate = test$estimate, 
                       p.value  = test$p.value
                       )
                   })
@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). For
instance, with 9 atlases and using \ants for registration and majority vote
fusion, the mean DSC scores for 1, 5, 9 and 17 templates are \Sexpr{k.1t},
\Sexpr{k.5t}, \Sexpr{k.9t}, \Sexpr{k.17t}, respectively.  A maximum
similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion.

The \ants{} registration method consistently outperforms \animal{} registration
over all variable settings we tested (mean increase in DSC is
\Sexpr{mean(ants$k.mb-animal$k.mb)}).  Pearson correlations of \mb{} DSC scores
when using weighted voting and when using non-weighted majority vote label
fusion (with \ants{} registration) for all combinations of atlases and templates
are $r > \Sexpr{min(ants.stats$estimate)}$,
\Sexpr{plevel(max(ants.stats$p.value))}, with a mean difference in DSC score of
\Sexpr{mean(mivote$k.mb - mavote$k.mb)}.  This result suggests that using a
weighted voting strategy does not significantly improve \mb{} segmentation
agreement, contrary to the findings of \citet{Aljabar2009} for basic
\multiatlas{} segmentation. Thus, in the remainder of our experiments only
results using the \ants{} registration algorithm and majority vote fusion will
be shown.

With at least five templates, \mb{} consistently shows a higher DSC score than
\multiatlas{} segmentation wth the same number of atlases: 
  $r = \Sexpr{round(cor.test(ants$k.mb, ants$k.ma)$estimate, 2)}$, 
       $\Sexpr{plevel(cor.test(ants$k.mb, ants$k.ma)$p.value)}$, 
mean DSC increase = \Sexpr{mean(ants$k.mb - ants$k.ma)} (Figure
\ref{fig:ADNI1-xval-k-diff}). The magnitude of DSC increase grows with template
library size but shows diminishing returns with larger atlas libraries.  Peak
increase (+\Sexpr{mb.diff.best} DSC) is found with a single atlas and template
library of 19 images.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expadni{}: \expadnidesc{}}
\label{exp:adni}
To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, we apply \mb{} to a large dataset
from the ADNI project. The resulting segmentations are compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated whole
hippocampal segmentations (\snt{}) provided by ADNI.

@@

\paragraph{\adnidataset{} dataset}  %%%%%%%%%%%
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper{}, and \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).  See Section
\ref{adnidataset} for study details, inclusion criteria and imaging
characteristics.

@@

@@

Figure \ref{fig:WAxval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subfield across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subfields are shown in Table
\ref{tab:WAxval-perturb-kappa}, along with DSC scores for the resampled atlas
segmentations when perturbed slightly and compared to the originals. We find
that the CA4/DG subfield shows the highest mean DSC score of
\Sexpr{mean(CA4DG)}$\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subfields having scores of \Sexpr{mean(CA1)}$\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)}$\pm$\Sexpr{sd(Sub)}, respectively. Both the CA4/DG and
molecular regions score below 0.5. These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion).
The whole hippocampus is segmented with a mean DSC score
of\Sexpr{mean(Whole)}$\pm$\Sexpr{sd(Whole)}.
